---
title: "Spatial Epi HW 3"
author: "David"
date: "2/16/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r install_packages, include=FALSE}

if (!isTRUE(requireNamespace("INLA", quietly = TRUE))) {
  install.packages("INLA", repos = c(getOption("repos"), 
          INLA = "https://inla.r-inla-download.org/R/stable"),
          dep=TRUE)
}


if (!require(SUMMER)) install.packages("SUMMER", repos = "http://cran.us.r-project.org")
if (!require(foreign)) install.packages("foreign", repos = "http://cran.us.r-project.org")
if (!require(haven)) install.packages("haven")
if (!require(rgeos)) install.packages("rgeos")
if (!require(rgdal)) install.packages("rgdal")
if (!require(maptools)) install.packages("maptools")
if (!require(sp)) install.packages("sp")
if (!require(spdep)) install.packages("spdep")
if (!require(SpatialEpi)) install.packages("SpatialEpi")
if (!require(RColorBrewer)) install.packages("RColorBrewer")
if (!require(ggplot2)) install.packages("ggplot2")
if (!require(maps)) install.packages("maps")
if (!require(broom)) install.packages("broom")
if (!require(raster)) install.packages("raster")
if (!require(leaflet)) install.packages("leaflet")
if (!require(dplyr)) install.packages("dplyr")
if (!require(tidyverse)) install.packages("tidyverse")
if (!require(SUMMER)) install.packages("SUMMER")


```


```{r load_packages, include=FALSE}

rm(list=ls())
library(foreign)
library(haven)
library(INLA)
library(rgdal)
library(maptools)
library(sp)
library(spdep)
library(SpatialEpi)
library(RColorBrewer)
library(ggplot2)
library(maps)
library(broom)
library(raster)
library(leaflet)
library(dplyr)
library(tidyverse)
library(SUMMER)
library(rgeos)

```


```{r, eval=TRUE, results='hide', warning=FALSE, message=FALSE, include=FALSE}

#Cancer data
#link = "https://github.com/dmccoomes/Spatial_epi/raw/master/HW%203/Data/ohio_data_ascending_fips.txt"
#ohio_canc <- read.table(url(link), header=TRUE)

#Cancer data
#on pc at home
#ohio_canc <- read.table("/Users/david/Documents/GitHub/Spatial_epi/HW 3/Data/ohio_2019_version.txt", header=TRUE)
#on laptop
#ohio_canc <- read.table("/Users/david/Documents/GitHub/Spatial_epi/HW 3/Data/ohio_2019_version.txt", header=TRUE)

#using github
link = "https://github.com/dmccoomes/Spatial_epi/raw/master/HW%203/Data/ohio_2019_version.txt"
ohio_canc <- read.table(url(link), header=TRUE)

#Map data
zip_oh_map <- "https://github.com/dmccoomes/Spatial_epi/raw/master/HW%203/Map%20data.zip"

#library(utils)
#temp=tempfile()
#download.file(zip_oh_map, temp)
#unzip(temp)

#(maps=list.files(pattern = 'shp'))

#ohmap <- readOGR(url(zip_oh_map),stringsAsFactors=F) 

#maplink <- "https://github.com/dmccoomes/Spatial_epi/raw/master/HW%203/Map%20data/"
#ohmap <- readOGR(dsn = maplink, layer="ohio_map")

#pc in office
ohmap <- readOGR(dsn="C:\\Users\\dcoomes\\Dropbox\\Classes\\Spatial modeling\\HW 3\\Data\\Map data", layer="ohio_map")

#laptop
#ohmap <- readOGR(dsn="/Users/david/Documents/GitHub/Spatial_epi/HW 3/Map data", layer="ohio_map")

#computer in library
#ohmap <- readOGR(dsn="C:\\Users\\dcoomes\\Documents\\GitHub\\Spatial_epi\\HW 3\\Map data", layer="ohio_map")

#ordering of regions is not the same among the data sets - how do we align these two?
#summary(ohmap)
#ohmap$COUNTYFP00

#summary(ohio_canc)
#ohio_canc$fips
#View(ohio_canc)


```

```{r create_smr, include=FALSE}

#creating smr
ohio_canc$smr <- ohio_canc$Obs/ohio_canc$Exp
#ohmap <- readOGR(dsn="C:\\Users\\dcoomes\\Documents\\GitHub\\Spatial_epi\\HW 3\\Data\\Map data", #layer="ohio_map")

#When using PC in office
#ohmap <- readOGR(dsn="C:\\Users\\dcoomes\\Desktop\\Epi PhD\\Classes\\Spatial epi\\Map data", layer="ohio_map")

#ordering of regions is not the same among the data sets - how do we align these two?
summary(ohmap)
ohmap$CNTYIDFP00

#summary(ohio_canc)
#ohio_canc$fips
#View(ohio_canc)

#creating RR
ohio_canc$Z <- log(ohio_canc$smr)
ohio_canc$varZ <- 1/(ohio_canc*ohio_canc$smr)
ohio_canc$precZ <- 1/ohio_canc$varZ

```


```{r merging_data, include=FALSE}

ohmap$fips <- as.numeric(as.character(ohmap$CNTYIDFP00))
#ohmap <- ohmap[order(ohmap$fips),]            #ordering fips so that counties align in data sets
ohmap <- merge(ohmap,ohio_canc,by="fips")

```



```{r inla_graph_file, include=FALSE}

#creating graph file (what is this doing?)
nb.map <- poly2nb(ohmap)
nb2INLA("ohmap.graph", nb.map)

#creating region variable in the cancer data
#ohio_canc <- ohio_canc[order(ohio_canc$fips),]     #not sure if I need to do this?

# I did this in reference to the map data, but I think that I should have done it in reference to the cancer data
ohio_canc$Region <- 1:nrow(ohio_canc)

#I get twice the number of regions as there are (176 as opposed to 88)
#ohmap$Region <- ohmap$fips

```

###Question 1
(a) The posterior median of $\beta$ is -0.0292 (95% CI: -0.0784-0.0138). The total variance of the random effects is 0.128 (95% CI: 0.082-0.128) and the proportion of the total variance that we can attribute to the spatial random effects is 0.0747 (95% CI: 0.0045-0.4967). 

```{r inla_fit, include=FALSE}

#for this we are fitting a spatially smoothed ICAR model using the INLA function

#Don't need to define pc.prec for this question
#pc.prec <- list(theta = list(prior = "pc.prec",
 #               param = c(1, 0.05)))

#ohmap2 <- as.data.frame(ohmap)

#setting up the inla model
formula <- Obs ~ 1 +    
        f(Region, model="bym2", graph="ohmap.graph", scale.model=T, constr=T,
          hyper=list(phi=list(prior="pc",
          param=c(0.5, 0.5), initial=1), prec=list(prior="pc.prec", 
          param=c(0.3,0.01), initial=5))) 

#fit the inla model
ohio.fit1 <- inla(formula, data=ohio_canc, family="poisson", 
                  E=Exp, 
                  control.predictor=list(compute=TRUE))

#ohio.fit1 <- inla(formula, data=ohio_canc, 
 #              family="gaussian",
  #             control.predictor=list(compute=TRUE),
   #            control.family = list(hyper = list(prec = list(initial = log(1), fixed=TRUE))),scale=precZ)


#get estimates
summary(ohio.fit1)
post.med.1 <- ohio.fit1$summary.fixed[4]              #posterior med
post.med.1_low <- ohio.fit1$summary.fixed[3]           #posterior med low   
post.med.1_high <- ohio.fit1$summary.fixed[5]          #posterior med high


var.ran <- 1/ohio.fit1$summary.hyperpar$`0.5quant`[1]        #total variance of random effects
var.ran_low <- 1/ohio.fit1$summary.hyperpar$`0.025quant`[1]
var.ran_high <- 1/ohio.fit1$summary.hyperpar$`0.975quant`[1]

var.ran

prop.1 <- ohio.fit1$summary.hyperpar$`0.5quant`[2]         #proportion of total variance due to random effects




```

(b) 

**Figure 1: Map of relative risk estimates for cancer in Ohio using an ICAR spatially smoothed model.** 

```{r map_setup_ICAR_1}

#diff is the non-spatial random effects
diff <- ohio.fit1$summary.random$Region[1:88, 2] - 
         ohio.fit1$summary.random$Region[89:176, 2]

REsnonspat <- exp(diff)
REsspat <- exp(ohio.fit1$summary.random$Region[89:176, 5])

ohmap$REsnonspat <- REsnonspat
ohmap$REsspat <- REsspat

#generating RR estimates
ohmap$postRR <- ohio.fit1$summary.fitted.values$`0.5quant`

spplot(ohmap, c("postRR"), col.regions=colorRampPalette(rev(brewer.pal(8, "RdBu")))(50))

#spplot(ohmap, c("REsnonspat"), col.regions=colorRampPalette(rev(brewer.pal(8, "RdBu")))(50))
#spplot(ohmap, c("REsspat"), col.regions=colorRampPalette(rev(brewer.pal(8, "RdBu")))(50))

```


<!--
```{r inla_nonspatial_map, include=FALSE}

ohio.fit1$summary.fixed

lnorminter.1 <- ohio.fit1$summary.fixed[4]
lnormREs.1 <- exp(ohio.fit1$summary.random$Region[5])

lnormRRs.2 <- as.double(exp(lnorminter.2)) * lnormREs.2[,1]


```



```{r inla_nonspatial_map2, include=FALSE}

ohmap$RRlnorm.2 <- lnormRRs.2
spplot(ohmap, c("RRlnorm.2"), col.regions=colorRampPalette(rev(brewer.pal(8, "RdBu")))(50))


```

-->


```{r inla_nonspatial_fit, include=FALSE}

#setting priors
pcprec <- list(theta = list(prior = "pc.prec",
          param = c(1, 0.05)))

#fit non-spatial map
ohio.fit2 <- inla(Obs ~ 1 + f(Region,
          model = "iid", hyper= pcprec), data = ohio_canc,
          family="poisson", E = Exp, control.predictor = list(compute=TRUE))

summary(ohio.fit2)

expbetaOmed <- ohio.fit2$summary.fixed[4]            #post median
sdmed <- 1/sqrt(ohio.fit2$summary.hyperpar[4])       #post sd

expbetaOmed
sdmed


```



```{r inla_nonspatial_map, include=FALSE}

ohio.fit2$summary.fixed

lnorminter.2 <- ohio.fit2$summary.fixed[4]
lnormREs.2 <- exp(ohio.fit2$summary.random$Region[5])

lnormRRs.2 <- as.double(exp(lnorminter.2)) * lnormREs.2[,1]


```


```{r inla_nonspatial_map2, include=FALSE}

ohmap$RRlnorm.2 <- lnormRRs.2
spplot(ohmap, c("RRlnorm.2"), col.regions=colorRampPalette(rev(brewer.pal(8, "RdBu")))(50))


```



```{r map_smr, include=FALSE}

spplot(ohmap, c("smr"), at = c(0,0.2,0.4,0.6, 0.8, 1, 1.2, 1.4, 1.6), col.regions = rev(brewer.pal(8, "RdBu")))

```


```{r plot_diff, include=FALSE}

plot(ohmap$smr, ohio.fit1$summary.fitted.values[,4],
     xlab="Gamma fitted", ylab="Spatial fitted")
abline(0,1)

```

**Figure 2: Plot of direct estimates (SMR) and spatially fitted estimates**

```{r compare_fit_smr}


plot(ohio_canc$smr, ohmap$postRR,
     xlab="Direct estimates (SMR)", ylab="Spatial fitted")
abline(0,1)


```


**Figure 3: Plot of non-spatially fitted estimates and spatially fitted estimates**

```{r compare_fit_nonspatial}

plot(ohmap$RRlnorm.2, ohmap$postRR, 
     xlab="Non-spatial fitted", ylab="Spatial fitted")
abline(0,1)

```



###Question 2


```{r load_kc_data, include=FALSE}

#reading in BRFSS data
data(BRFSS)
#summary(BRFSS)

#removing obs with missing hracode and smoking status
brfss <- subset(BRFSS, !is.na(BRFSS$hracode))
brfss <- subset(brfss, !is.na(brfss$smoker1))

#load spatial data
data(KingCounty)
nb.r <- poly2nb(KingCounty, queen=F,
                row.names = KingCounty$HRA2010v2_)
mat <- nb2mat(nb.r, style="B", zero.policy = TRUE)
colnames(mat) <- rownames(mat)
mat <- as.matrix(mat[1:dim(mat)[1], 1:dim(mat)[1]])
mat[1:2, 1:2]

```


```{r survey, include=FALSE}

#calculating weighted non-smoothed estimates
library(survey)

#creating survey weights
design <- svydesign(ids=~1, weights = ~rwt_llcp,
                    strata = ~strata, data=brfss)
direct <- svyby(~smoker1, ~hracode, design, svymean)
head(direct, n=2)

```


```{r generic_fit, include=FALSE}

#smoothed, non-weighted estimates and non-smoothed, non-weighted estimates
smoothed <- fitGeneric(data=brfss, geo=KingCounty,
      Amat=mat, responseType = "binary", responseVar="smoker1",
      strataVar=NULL, weightVar=NULL, regionVar="hracode",
      clusterVar=NULL, CI=0.95)
head(smoothed$HT, n=2)

#The smooth estimates are the smoothed non-weighted estimates and the HT estimates are the non-smoothed, non-weighted estimates
head(smoothed$smooth, n=1)
#simple smoothed binomial probabilities
head(smoothed$HT, n=1)



```


```{r map_smooth, include=FALSE}

if (!require(gpclib)) install.packages("gpclib", type="source")
gpclibPermit()

#mapping smoothed, non-weighted estimates
toplot <- smoothed$smooth
mapPlot(data = toplot, geo=KingCounty,
        variables=c("mean.original"), 
        labels=c("Posterior Mean"), by.data="region",
        by.geo="HRA2010v2_")

```


```{r map_weight_smooth}


svysmoothed <- fitGeneric(data=brfss, geo=KingCounty,
       Amat = mat, responseType="binary", responseVar="smoker1",
       strataVar="strata", weightVar="rwt_llcp", regionVar="hracode",
       clusterVar="~1", CI=0.95)


est <- data.frame(naive=smoothed$HT$HT.est.original,
    weighted=svysmoothed$HT$HT.est.original, smooth=smoothed$smooth$mean.original,
    weightedsmooth=svysmoothed$smooth$mean.original)

var <- data.frame(naive=smoothed$HT$HT.variance.original,
    weighted=svysmoothed$HT$HT.variance.original,
    smooth=smoothed$smooth$variance.original, weightedsmooth=svysmoothed$smooth)

```


```{r model_output, include=FALSE}

P2 <- data.frame(HRA=smoothed$HT$region, "Simple Proportion"=smoothed$HT$HT.est.original, "Posterior Median Estimates"=smoothed$smooth$median.original, "Weighted Direct Estimates"=svysmoothed$HT$HT.est.original, 
"Smoothed Direct Estimates"=svysmoothed$smooth$median.original)

P2[1:48,]

```



```{r model_compare}

est <- data.frame(naive = smoothed$HT$HT.est.original,
                  weighted = svysmoothed$HT$HT.est.original,
                  smooth = smoothed$smooth$mean.original,
                  weightedsmooth = svysmoothed$smooth$mean.original)

l1 <- range(est)

g1 <- ggplot(est, aes(x = naive, y = smooth)) + geom_point() + geom_abline(slope = 1, intercept = 0, color = "red") + geom_smooth(method = "lm") + ggtitle("Unweighted") + xlab("Unsmoothed") + ylab("Smoothed")+ xlim(l1) + ylim(l1) 
g2 <- ggplot(est, aes(x = naive, y = weighted)) + geom_point() + geom_abline(slope = 1, intercept = 0, color = "red") + geom_smooth(method = "lm") + ggtitle("Unsmoothed") + xlab("Unweighted") + ylab("Weighted")+ xlim(l1) + ylim(l1) 
g3 <- ggplot(est, aes(x = weighted, y = weightedsmooth)) + geom_point() + geom_abline(slope = 1, intercept = 0, color = "red") + geom_smooth(method = "lm") + ggtitle("Weighted") + xlab("Unsmoothed") + ylab("Smoothed") + xlim(l1) + ylim(l1)
g4 <- ggplot(est, aes(x = smooth, y = weightedsmooth)) + geom_point() + geom_abline(slope = 1, intercept = 0, color = "red") + geom_smooth(method = "lm") + ggtitle("Smoothed") + xlab("Unweighted") + ylab("Weighted")+ xlim(l1) + ylim(l1) 
print(g1)
print(g2)
print(g3)
print(g4)
library(gridExtra)
grid.arrange(grobs = list(g1, g3, g2, g4), ncol = 2)


```



