---
title: "Spatial Epi HW 3"
author: "David"
date: "2/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r install_packages, include=FALSE}

if (!isTRUE(requireNamespace("INLA", quietly = TRUE))) {
  install.packages("INLA", repos = c(getOption("repos"), 
          INLA = "https://inla.r-inla-download.org/R/stable"),
          dep=TRUE)
}


if (!require(SUMMER)) install.packages("SUMMER", repos = "http://cran.us.r-project.org")
if (!require(foreign)) install.packages("foreign", repos = "http://cran.us.r-project.org")
if (!require(haven)) install.packages("haven")
if (!require(rgdal)) install.packages("rgdal")
if (!require(maptools)) install.packages("maptools")
if (!require(sp)) install.packages("sp")
if (!require(spdep)) install.packages("spdep")
if (!require(SpatialEpi)) install.packages("SpatialEpi")
if (!require(RColorBrewer)) install.packages("RColorBrewer")
if (!require(ggplot2)) install.packages("ggplot2")
if (!require(maps)) install.packages("maps")
if (!require(broom)) install.packages("broom")
if (!require(raster)) install.packages("raster")
if (!require(leaflet)) install.packages("leaflet")
if (!require(dplyr)) install.packages("dplyr")
if (!require(tidyverse)) install.packages("tidyverse")



#Can add more here if needed

```


```{r load_packages, include=FALSE}

rm(list=ls())
library(foreign)
library(haven)
#Having trouble installing INLA on pc in office
#install.packages("INLA", repos=c(getOption("repos"), INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
library(INLA)
library(rgdal)
library(maptools)
library(sp)
library(spdep)
library(SpatialEpi)
library(RColorBrewer)
library(ggplot2)
library(maps)
library(broom)
library(raster)
library(leaflet)
library(dplyr)
library(tidyverse)
library(SUMMER)

```


```{r, eval=TRUE, results='hide', warning=FALSE, message=FALSE, include=FALSE}

#Cancer data
#link = "https://github.com/dmccoomes/Spatial_epi/raw/master/HW%203/Data/ohio_data_ascending_fips.txt"
#ohio_canc <- read.table(url(link), header=TRUE)

#Cancer data
#ohio_canc <- read.table("/Users/david/Documents/GitHub/Spatial_epi/HW 3/Data/ohio_2019_version.txt", header=TRUE)

ohio_canc <- read.table("/Users/david/Documents/GitHub/Spatial_epi/HW 3/Data/ohio_2019_version.txt", header=TRUE)

#Map data
zip_oh_map <- "https://github.com/dmccoomes/Spatial_epi/raw/master/HW%203/Map%20data.zip"

#library(utils)
#temp=tempfile()
#download.file(zip_oh_map, temp)
#unzip(temp)

#(maps=list.files(pattern = 'shp'))

#ohmap <- readOGR(url(zip_oh_map),stringsAsFactors=F) 

ohmap <- readOGR(dsn="/Users/david/Documents/GitHub/Spatial_epi/HW 3/Map data", layer="ohio_map")

#ordering of regions is not the same among the data sets - how do we align these two?
#summary(ohmap)
#ohmap$COUNTYFP00

#summary(ohio_canc)
#ohio_canc$fips
#View(ohio_canc)


```

```{r create_smr, include=FALSE}

#creating smr
ohio_canc$smr <- ohio_canc$Obs/ohio_canc$Exp

#creating RR
ohio_canc$Z <- log(ohio_canc$smr)
ohio_canc$varZ <- 1/(ohio_canc*ohio_canc$smr)
ohio_canc$precZ <- 1/ohio_canc$varZ

```


```{r merging_data, include=FALSE}

ohmap$fips <- as.numeric(as.character(ohmap$CNTYIDFP00))
ohmap <- ohmap[order(ohmap$fips),]            #ordering fips so that counties align in data sets
ohmap <- merge(ohmap,ohio_canc,by="fips")

```



```{r inla_graph_file, include=FALSE}

#creating graph file (what is this doing?)
nb.map <- poly2nb(ohmap)
nb2INLA("ohmap.graph", nb.map)

#creating region variable in the cancer data
#ohio_canc <- ohio_canc[order(ohio_canc$fips),]     #not sure if I need to do this?
ohio_canc$Region <- 1:nrow(ohmap)

#I get twice the number of regions as there are (176 as opposed to 88)
#ohmap$Region <- ohmap$fips

```

###Question 1
(a) The posterior median of $\beta$ is -0.0292 (95% CI: -0.279-0.221). The total variance of the random effects is 0.0163 and the proportion of the total variance that we can attribute to the spatial random effects is 0.0747. 

```{r inla_fit, include=FALSE}

#for this we are fitting a spatially smoothed ICAR model using the INLA function

#Don't need to define pc.prec for this question
#pc.prec <- list(theta = list(prior = "pc.prec",
 #               param = c(1, 0.05)))

#ohmap2 <- as.data.frame(ohmap)

#setting up the inla model
formula <- Obs ~ 1 +    
        f(Region, model="bym2", graph="ohmap.graph", scale.model=T, constr=T,
          hyper=list(phi=list(prior="pc",
          param=c(0.5, 0.5), initial=1), prec=list(prior="pc.prec", 
          param=c(0.3,0.01), initial=5))) 

#fit the inla model
ohio.fit1 <- inla(formula, data=ohio_canc, family="poisson", 
                  E=Exp, 
                  control.predictor=list(compute=TRUE))


#ohio.fit1 <- inla(formula, data=ohio_canc, 
 #              family="gaussian",
  #             control.predictor=list(compute=TRUE),
   #            control.family = list(hyper = list(prec = list(initial = log(1), fixed=TRUE))),scale=precZ)


#get estimates
summary(ohio.fit1)
post.med.1 <- ohio.fit1$summary.fixed[4]                       #posterior med
sdmed.1 <- 1/sqrt(ohio.fit1$summary.hyperpar$`0.5quant`[1])     #posterior sd
beta_low <- post.med.1 - 1.96*sdmed.1
beta_high <- post.med.1 + 1.96*sdmed.1

re.var <- 1/ohio.fit1$summary.hyperpar$`0.5quant`[1]      #variance of the random effects
prop.spa <- ohio.fit1$summary.hyperpar$`0.5quant`[2]      #proportion of random effects due to spatial


```

(b) 

**Figure 1: Map of relative risk estiamtes for cancer in Ohio using ICAR spatially smoothed model.** 

```{r map_setup_ICAR_1}

diff <- ohio.fit1$summary.random$Region[1:88, 2] - 
         ohio.fit1$summary.random$Region[89:176, 2]

REsnonspat <- exp(diff)
REsspat <- exp(ohio.fit1$summary.random$Region[89:176, 5])

ohmap$REsnonspat <- REsnonspat
ohmap$REsspat <- REsspat
spplot(ohmap, c("REsnonspat"), col.regions=colorRampPalette(rev(brewer.pal(8, "RdBu")))(50))

```


```{r inla_nonspatial_fit, include=FALSE}

#setting priors
pcprec <- list(theta = list(prior = "pc.prec",
          param = c(1, 0.05)))

#fit non-spatial map
ohio.fit2 <- inla(Obs ~ 1 + f(Region,
          model = "iid", hyper= pcprec), data = ohio_canc,
          family="poisson", E = Exp, control.predictor = list(compute=TRUE))

summary(ohio.fit2)

expbetaOmed <- ohio.fit2$summary.fixed[4]            #post median
sdmed <- 1/sqrt(ohio.fit2$summary.hyperpar[4])       #post sd

expbetaOmed
sdmed


```


**Figure 2: Map of relative risk of cancer in Ohio using lognormal non-spatially smoothed model.**

```{r inla_nonspatial_map, include=FALSE}

ohio.fit2$summary.fixed

lnorminter.2 <- ohio.fit2$summary.fixed[4]
lnormREs.2 <- exp(ohio.fit2$summary.random$Region[5])

lnormRRs.2 <- as.double(exp(lnorminter.2)) * lnormREs.2[,1]


```


```{r inla_nonspatial_map2}

ohmap$RRlnorm.2 <- lnormRRs.2
spplot(ohmap, c("RRlnorm.2"), col.regions=colorRampPalette(rev(brewer.pal(8, "RdBu")))(50))


```


**Figure 3: Map of SMR of cancer in Ohio.**

```{r map_smr}

spplot(ohmap, c("smr"), at = c(0,0.2,0.4,0.6, 0.8, 1, 1.2, 1.4, 1.6), col.regions = rev(brewer.pal(8, "RdBu")))

```


```{r plot_diff, include=FALSE}

plot(ohmap$smr, ohio.fit1$summary.fitted.values[,4],
     xlab="Gamma fitted", ylab="Spatial fitted")
abline(0,1)

```





```{r load_kc_data, include=FALSE}

#reading in BRFSS data
data(BRFSS)
#summary(BRFSS)

#removing obs with missing hracode and smoking status
brfss <- subset(BRFSS, !is.na(BRFSS$hracode))
brfss <- subset(brfss, !is.na(brfss$smoker1))

#load spatial data
data(KingCounty)
nb.r <- poly2nb(KingCounty, queen=F,
                row.names = KingCounty$HRA2010v2_)
mat <- nb2mat(nb.r, style="B", zero.policy = TRUE)
colnames(mat) <- rownames(mat)
mat <- as.matrix(mat[1:dim(mat)[1], 1:dim(mat)[1]])
mat[1:2, 1:2]

```


```{r survey, include=FALSE}

#calculating weighted non-smoothed estimates
library(survey)
design <- svydesign(ids=~1, weights = ~rwt_llcp,
                    strata = ~strata, data=brfss)
direct <- svyby(~smoker1, ~hracode, design, svymean)
head(direct, n=2)

```


```{r generic_fit, include=FALSE}

#smoothed, non-weighted estimates
smoothed <- fitGeneric(data=brfss, geo=KingCounty,
      Amat=mat, responseType = "binary", responseVar="smoker1",
      strataVar=NULL, weightVar=NULL, regionVar="hracode",
      clusterVar=NULL, CI=0.95)
head(smoothed$HT, n=2)

#
head(smoothed$smooth, n=1)
#simple smoothed binomial probabilities
head(smoothed$HT, n=1)



```


```{r map_smooth_nw}

#mapping smoothed, non-weighted estimates
toplot <- smoothed$smooth
mapPlot(data = toplot, geo=KingCounty,
        variables=c("mean.original"), 
        labels=c("Posterior Mean"), by.data="region",
        by.geo="HRA2010v2_")

```

