---
title: "Spatial Epi - Final Project"
author: "David Coomes"
date: "January 27, 2020"
output: html_document
---

### 1. Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
```


```{r install_packages, include=FALSE}

if (!isTRUE(requireNamespace("INLA", quietly = TRUE))) {
  install.packages("INLA", repos = c(getOption("repos"), 
          INLA = "https://inla.r=inla-download.org/R/stable"),
          dep=TRUE)
}


if (!require(SUMMER)) install.packages("SUMMER", repos = "http://cran.us.r-project.org")
if (!require(foreign)) install.packages("foreign", repos = "http://cran.us.r-project.org")
if (!require(zscorer)) install.packages("zscorer")
if (!require(naniar)) install.packages("naniar")
if (!require(anthro)) install.packages("anthro")
if (!require(tmaptools)) install.packages("tmaptools")
if (!require(rmapshaper)) install.packages("rmapshaper")
if (!require(tmap)) install.packages("tmap")
if (!require(rmapshaper)) install.packages("rmapshaper")
if (!require(doBy)) install.packages("doBy")


```


```{r load_packages, include=FALSE}

rm(list=ls())
library(foreign)
library(haven)
#Having trouble installing INLA on pc in office
#install.packages("INLA", repos=c(getOption("repos"), INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
library(INLA)
library(rgdal)
library(maptools)
library(sp)
library(spdep)
library(SpatialEpi)
library(RColorBrewer)
library(ggplot2)
library(maps)
library(broom)
library(raster)
library(leaflet)
library(dplyr)
library(tidyverse)
library(SUMMER)
library(zscorer)
library(naniar)

```


### 2. Data cleaning

Starting with Ethiopia Wave 3

```{r load_EPAR_data}

#loading health data
link_3 = "https://github.com/dmccoomes/Spatial_epi/raw/master/Final%20project/Raw%20data/ESS%20Wave%203/Ethiopia%20Wave%203%20data/ETH_2015_ESS_v02_M_STATA8/Household/sect3_hh_w3.dta"

et_w3_anth <- read_dta(url(link_3))
# head(et_w3_anth)
# summary(et_w3_anth)
# View(et_w3_anth)

#keeping only those we need
keep_vars.anth <- c("household_id2", "rural", "pw_w3", "individual_id2", "ea_id2", "saq01", "saq02", "saq03", "saq04", "saq05", "saq06", "saq07", "hh_s3q12_a", "hh_s3q20a", "hh_s3q20d", "hh_s3q20e", "hh_s3q21_a", "hh_s3q21_b", "hh_s3q21_c", "hh_s3q22", "hh_s3q23")    
et_w3_anth <- et_w3_anth[keep_vars.anth]

#renaming variables
et_w3_anth <- et_w3_anth %>%
  rename(
    u_5 = hh_s3q12_a,
    diarrhea_2wk = hh_s3q20a,
    breastfed_yn = hh_s3q20e,
    birth_day = hh_s3q21_a,
    birth_month = hh_s3q21_b,
    birth_year = hh_s3q21_c,
    weight_kg = hh_s3q22,
    height_cm = hh_s3q23
  )
  
#downloading cover sheet data
link_cover = "https://github.com/dmccoomes/Spatial_epi/raw/master/Final%20project/Raw%20data/ESS%20Wave%203/Ethiopia%20Wave%203%20data/ETH_2015_ESS_v02_M_STATA8/Household/sect_cover_hh_w3.dta"

et_w3_cover <- read_dta(url(link_cover))
# View(et_w3_cover)

et_w3_cover <- et_w3_cover %>%
  rename(
    int_day=hh_saq13_a,
    int_month=hh_saq13_b,
    int_year=hh_saq13_c
  )

#downloading section 1
link_1 = "https://github.com/dmccoomes/Spatial_epi/raw/master/Final%20project/Raw%20data/ESS%20Wave%203/Ethiopia%20Wave%203%20data/ETH_2015_ESS_v02_M_STATA8/Household/sect1_hh_w3.dta"
et_w3_hh <- read_dta(url(link_1))

#get number in HH
et_w3_hh$hh_member <- 1
et_w3_hh_2 <- et_w3_hh %>%
  group_by(household_id2) %>%
    summarize(hh_members = sum(hh_member))

et_w3_hh_2 <- et_w3_hh_2[c("household_id2", "hh_members")]

keep_vars.hh <- c("household_id2", "rural", "pw_w3", "individual_id2", "ea_id2", "saq01", "saq02", "saq03", "saq04", "saq05", "saq06", "saq07", "hh_s1q03", "hh_s1q04a", "hh_s1q04b", "hh_s1q04h", "saq08")    
et_w3_hh <- et_w3_hh[keep_vars.hh]


#Merging household cover data with health data
#et_w3_health <- merge(et_w3_anth, et_w3_cover, no.dups=TRUE)    #I don't think we need the cover anymore
#View(et_w3_health)
 
et_w3_hh <- merge(et_w3_hh, et_w3_hh_2, no.dups=TRUE)
et_w3_health <- merge(et_w3_anth, et_w3_hh, no.dups=TRUE)

#merge in geovariables
link_geovars = "https://github.com/dmccoomes/Spatial_epi/raw/master/Final%20project/Raw%20data/ESS%20Wave%203/Ethiopia%20Wave%203%20data/ETH_2015_ESS_v02_M_STATA8/Geovariables/ETH_HouseholdGeovars_y3.dta"
et_w3_geovars <- read_dta(url(link_geovars))

#merging with dataset
et_w3_health <- merge(et_w3_health, et_w3_geovars, no.dups=TRUE)

et_w3_health <- et_w3_health %>%
  rename(
    latitude=lat_dd_mod,
    longitude=lon_dd_mod
  )

```



```{r create_vars}

#keeping only some variables
#keep_vars <- c("household_id2", "rural", "pw_w3", "individual_id2", "ea_id2", "saq01")

#generating age in months
et_w3_health$age_months <- et_w3_health$hh_s1q04a*12 + et_w3_health$hh_s1q04b

#subset - only keep children under 5
#et_w3_data <- subset(et_w3_health, u_5==1)
et_w3_health$female <- et_w3_health$hh_s1q03
et_w3_data <- subset(et_w3_health, age_months <=60)
#recoding sex
et_w3_health[,"female"] <- ifelse(et_w3_health[,"female"]==1,0, ifelse(et_w3_health[,"female"]==2,1, 99))


#renaming variables
et_w3_data <- et_w3_data %>%
  rename(
    age=hh_s1q04h,
    region=saq01,
    zone=saq02,
    woreda=saq03,
    town=saq04,
    subcity=saq05,
    kebele=saq06,
    ea=saq07,
    hh=saq08
  )

#generating age in days
et_w3_data$age_days <- et_w3_data$age_months*30

et_w3_data <- et_w3_data  %>%
  select(household_id2, rural, pw_w3, individual_id2, ea_id2, region, zone, woreda, town, subcity, kebele, ea, hh, u_5, weight_kg, height_cm,
         birth_month, birth_year, age, age_months, female, hh_s1q03, age_days, latitude, longitude, hh_members)

#summary(et_w3_data$age)

# colnames(et_w3_data)
# summary(et_w3_data$age)

#keeping only those that report 

#Generating age using date of interview and year of birth
#et_w3_health$year <- et_w3_health$ et_w3_health$hh_saq13_c

#adding anthropometric data
#addWGSR(data=et_w3_data, sex="hh_s1q03", firstPart="weight_kg", secondPart="height_cm", thirdPart="age_days", index="wfa", standing=NULL, output="wfa_z", digits=3)

#setting as null if height is not between 45-120 cm
#et_w3_health[,"height_cm"] <- ifelse(et_w3_health[,"height_cm"]<45,NA, ifelse(et_w3_health[,"height_cm"]>120, NA, et_w3_health[,"height_cm"]))
et_w3_data$height_cm[et_w3_data$height_cm<45] <- NA
et_w3_data$height_cm[et_w3_data$height_cm>120] <- NA
et_w3_data$age_months[et_w3_data$age_months>60] <- NA


# summary(et_w3_data$age_months) 
# summary(et_w3_data$height_cm)

# et_w3_data$waz <- getAllWGS(data=et_w3_data, sex="hh_s1q03", weight="weight_kg", height="height_cm", age="age_months", index="wfa")
# et_w3_data$haz <- getAllWGS(data=et_w3_data, sex="hh_s1q03", weight="weight_kg", height="height_cm", age="age_months", index="hfa")
# et_w3_data$whz <- getAllWGS(data=et_w3_data, sex="hh_s1q03", weight="weight_kg", height="height_cm", age="age_months", index="wfh")

#et_w3_data$all <- getAllWGS(data=et_w3_data, sex="hh_s1q03", weight="weight_kg", height="height_cm", age="age_months", index="all")

et_w3_data$waz <- getCohortWGS(data=et_w3_data,
                                 sexObserved = "hh_s1q03",
                                 firstPart = "weight_kg",
                                 secondPart = "age_months",
                                 index = "wfa")

et_w3_data$haz <- getCohortWGS(data=et_w3_data,
                                 sexObserved = "hh_s1q03", 
                                 firstPart = "height_cm",
                                 secondPart = "age_months",
                                 index = "hfa")

#this isn't working - using the anthro_zscores package below
# et_w3_data$whz <- getCohortWGS(data=et_w3_data,
#                                  sexObserved = "hh_s1q03",
#                                  firstPart = "weight_kg",
#                                  secondPart = "height_cm",
#                                  index = "wfh")
                            

library(anthro)

zscore <- with(et_w3_data,
     anthro_zscores(
        sex = hh_s1q03,
        age = age_months,
        is_age_in_month = TRUE,
        weight = weight_kg,
        lenhei = height_cm))

et_w3_data <- cbind(et_w3_data, zscore)

#renaming variables
et_w3_data <- et_w3_data %>%
  rename(
    HAZ = zlen,
    WAZ = zwei,
    WHZ = zwfl
  )

et_w3_data$stunted <- ifelse(et_w3_data$HAZ <= -2, 1, 0)
et_w3_data$wasted <- ifelse(et_w3_data$WHZ <= -2, 1, 0)

#generating individual weights
et_w3_data$indiv_wgt <- et_w3_data$pw_w3*et_w3_data$hh_members

```


Creating variables of interest
```{r create_variables, include=FALSE, eval=FALSE}

#everyone born from 1-15 days of month get that month
# et_w3_data$age_month <- (et_w3_data$int_year - et_w3_data$birth_year)*12 + (et_w3_data$int_month - et_w3_data$birth_month) 
# summary(et_w3_data$age_month)

```



Reading the shape file into R
```{r map, eval=TRUE, results='hide', warning=FALSE, message=FALSE}

#Bringing in maps
library(rgdal)
#on laptop
#et_admin2_map <- readOGR("/Users/david/Documents/GitHub/Spatial_epi/Final project/Map data/ETH_admin_SHP",stringsAsFactors=F) 

#pc in office
# et_admin2_map <- readOGR(dsn="C:\\Users\\dcoomes\\Dropbox\\Classes\\Spatial modeling\\Final project\\Map data\\ETH_admin_SHP", stringsAsFactors=F)
# et_admin1_map <- readOGR(dsn="C:\\Users\\dcoomes\\Dropbox\\Classes\\Spatial modeling\\Final project\\Map data\\ETH-level_1_SHP", stringsAsFactors=F)
# et_boundary_map <- readOGR(dsn="C:\\Users\\dcoomes\\Dropbox\\Classes\\Spatial modeling\\Final project\\Map data\\ETH_boundaries_SHP", stringsAsFactors=F)   #this looks like an admin 2 map

#new region and zone maps that match up with data
# et_admin1_map.3 <- readOGR("C:\\Users\\dcoomes\\Dropbox\\Classes\\Spatial modeling\\Final project\\Map data\\ethiopiaregion", stringsAsFactors = F)
# et_admin2_map.2 <- readOGR("C:\\Users\\dcoomes\\Dropbox\\Classes\\Spatial modeling\\Final project\\Map data\\ethiopia-zone", stringsAsFactors = F)


#home desktop
# et_admin1_map.2 <- readOGR("/Users/david/Documents/GitHub/Spatial_epi/Final project/Map data/ETH-level_1_SHP", stringsAsFactors = F)
# et_admin2_map <- readOGR("/Users/david/Documents/GitHub/Spatial_epi/Final project/Map data/ETH_admin_SHP", stringsAsFactors = F)

#new region and zone maps that match up with data
et_admin1_map.3 <- readOGR("/Users/david/Desktop/Dave/school/Spatial Epi/Final project/map data/ethiopiaregion", stringsAsFactors = F)
et_admin2_map.2 <- readOGR("/Users/david/Desktop/Dave/school/Spatial Epi/Final project/map data/ethiopia-zone", stringsAsFactors = F)

#changing utm to lat and long
#sputm <- SpatialPoints(randompoints, proj4string = CRS("+proj=utm +zone=37 +ellps=clrk80 +towgs84=-166,-15,204,0,0,0,0"))  #don't need this
et_admin1_map <- spTransform(et_admin1_map.3, CRS("+proj=longlat +datum=WGS84"))
et_admin2_map <- spTransform(et_admin2_map.2, CRS("+proj=longlat + datum=WGS84"))

```


```{r plot_map, include=FALSE}

#plot each map - boundary is the same as admin2
plot(et_admin2_map)
plot(et_admin1_map)
#plot(et_boundary_map)

#if you want to layer
plot(et_admin1_map)
plot(et_admin2_map, add=T)

```


```{r subset_data}

#starting with those that include HAZ estimates
et_w3_data <- subset(et_w3_data, !is.na(et_w3_data$HAZ))
#keeping only rural respondents
et_w3_data <- subset(et_w3_data, et_w3_data$rural==1)

#creating new region string variable
et_w3_data$region.string <- as.character(et_w3_data$region)
et_w3_data$zone.string <- as.character(et_w3_data$zone)

et_w3_data$region.string[et_w3_data$region.string=="1"] <- "01"
et_w3_data$region.string[et_w3_data$region.string=="2"] <- "02"
et_w3_data$region.string[et_w3_data$region.string=="3"] <- "03"
et_w3_data$region.string[et_w3_data$region.string=="4"] <- "04"
et_w3_data$region.string[et_w3_data$region.string=="5"] <- "05"
et_w3_data$region.string[et_w3_data$region.string=="6"] <- "06"
et_w3_data$region.string[et_w3_data$region.string=="7"] <- "07"

et_w3_data$zone.string[et_w3_data$zone.string=="1"] <- "01"
et_w3_data$zone.string[et_w3_data$zone.string=="2"] <- "02"
et_w3_data$zone.string[et_w3_data$zone.string=="3"] <- "03"
et_w3_data$zone.string[et_w3_data$zone.string=="4"] <- "04"
et_w3_data$zone.string[et_w3_data$zone.string=="5"] <- "05"
et_w3_data$zone.string[et_w3_data$zone.string=="6"] <- "06"
et_w3_data$zone.string[et_w3_data$zone.string=="7"] <- "07"
et_w3_data$zone.string[et_w3_data$zone.string=="8"] <- "08"
et_w3_data$zone.string[et_w3_data$zone.string=="9"] <- "09"

et_w3_data <- within(et_w3_data, zone.st <- paste(region.string, zone.string, sep=""))

#only keeping those with zone codes that exist in the map
et_w3_data_zone <- subset(et_w3_data, zone.st != "0106" | zone.st != "0311" | zone.st != "0415" | zone.st != "0416" | zone.st != "0606" | zone.st != "0714" | zone.st != "0715" | zone.st != "0716" | zone.st != "0722" | zone.st != "1204" | zone.st != "1402" | zone.st != "1403" | zone.st != "1404" | zone.st != "1405" | zone.st != "1406" | zone.st != "1407" | zone.st != "1408" | zone.st != "1409" | zone.st != "1410")


#developing stratas
et_w3_data$strata[et_w3_data$rural==1] <- et_w3_data$region                    #strata is region if rural
et_w3_data$strata[et_w3_data$rural==2 & et_w3_data$region == 1] <- 16
et_w3_data$strata[et_w3_data$rural==2 & et_w3_data$region == 3] <- 17
et_w3_data$strata[et_w3_data$rural==2 & et_w3_data$region == 4] <- 18
et_w3_data$strata[et_w3_data$rural==2 & et_w3_data$region == 7] <- 19
et_w3_data$strata[et_w3_data$rural==2 & (et_w3_data$region == 5 | et_w3_data$region == 6 | et_w3_data$region == 12 | et_w3_data$region == 13 | et_w3_data$region == 15)] <- 20
et_w3_data$strata[et_w3_data$rural==3 & et_w3_data$region == 1] <- 21
et_w3_data$strata[et_w3_data$rural==3 & et_w3_data$region == 3] <- 22
et_w3_data$strata[et_w3_data$rural==3 & et_w3_data$region == 4] <- 23
et_w3_data$strata[et_w3_data$rural==3 & et_w3_data$region == 7] <- 24
et_w3_data$strata[et_w3_data$rural==3 & et_w3_data$region == 14] <- 25
et_w3_data$strata[et_w3_data$rural==3 & (et_w3_data$region == 5 | et_w3_data$region == 6 | et_w3_data$region == 12 | et_w3_data$region == 13 | et_w3_data$region == 15)] <- 26


```


### 3. Set up spatial maps

```{r set_up_spatial_map, echo=FALSE}

map_desc <- as.data.frame(et_admin1_map)
map_desc.2 <- as.data.frame(et_admin2_map)

nb.r <- poly2nb(et_admin1_map, queen=F, row.names = et_admin1_map$REG_P_CODE)    #couldn't put rownames here - do I need those?
mat <- nb2mat(nb.r, style="B", zero.policy = TRUE)
colnames(mat) <- rownames(mat)
mat <- as.matrix(mat[1:dim(mat)[1], 1:dim(mat)[1]])

mat

nb.r.zone <- poly2nb(et_admin2_map, queen=F, row.names = et_admin2_map$ZON_P_CODE)
mat.2 <- nb2mat(nb.r.zone, style="B", zero.policy=TRUE)
colnames(mat.2) <- rownames(mat.2)
mat.2 <- as.matrix(mat.2[1:dim(mat.2)[1], 1:dim(mat.2)[1]])

mat.2

```


```{r check_clusters}

library(tmaptools)
library(raster)

mapCRS=crs(et_admin1_map) # projection of our map

contriET_geo <- SpatialPointsDataFrame(et_w3_data[,c("longitude", "latitude")], # Lon/LAT
                    et_w3_data,    #the original data frame
                    proj4string = mapCRS)   # assign a CRS of map 

get_proj4(mapCRS)

library(rmapshaper)
# This will make just a border of the state
base_admin1 <- ms_dissolve(et_admin1_map)


# plot(et_admin1_map,border='black',lwd=1)
# 
# # NEW LAYER ON TOP
# plot(contriET_geo, 
#      pch=20, #type of character (see link above)
#      col = "red", 
#      cex = .6, # shrinking
#      add=T) # VERY important

library(tmap)

creditsText="EPSG: 4326\nProj=longlat\ndatum=WGS84"

tm_shape(base_admin1) + tm_borders("black", lwd=2) + tm_polygons() +
    tm_shape(et_admin1_map) + tm_borders("black", lwd=1) +
    #ALPHA 0 is total transparency
    tm_shape(contriET_geo) + tm_dots(size = 0.1,col = 'red',alpha=0.5) + 
    tm_layout(main.title = "Ethiopia LSMS Cluster locations",main.title.position = 'center') +
    tm_compass(position = c('left','TOP'),type = 'arrow') 
    #tm_scale_bar(position=c("RIGHT", "BOTTOM"),width = 0.2) +
    #tm_credits(creditsText, position=c("left", "bottom"))


```




```{r survey_design}

#getting regional means
library(survey)
design <- svydesign(ids = ~1, weights = ~indiv_wgt,
                    strata = ~strata, nest=TRUE, cluster= ~ea_id2, data=et_w3_data)

direct_region <- svyby(~stunted, ~strata + region, design, svymean)
head(direct_region)

direct_zone <- svyby(~stunted, ~region + zone, design, svymean)
head(direct_zone)


```


Why is this model not fitting? Can I not fit any generic model with my data? It doesn't seem to matter how I change the variables in the model

```{r fit_icar}

et_w3_data_sub <- subset(et_w3_data, !is.na(indiv_wgt) & !is.na(stunted) & !is.na(region) & !is.na(ea_id2))

svysmoothed <- fitGeneric(data=et_w3_data_sub, geo=contriET_geo,
                       Amat = mat, responseType="binary", responseVar="stunted",
                       strataVar = NULL, weightVar = "indiv_wgt", regionVar = "region",
                       clusterVar = "~ea_id2", CI=0.95)


# svysmoothed <- fitGeneric(data=et_w3_data, geo=et_admin1_map,
#                        Amat = mat, responseType="binary", responseVar="stunted",
#                        strataVar = NULL, weightVar = "indiv_wgt", regionVar = "region",
#                        clusterVar = NULL, CI=0.95)


```


```{r map_smoothed_est}

library(mapproj)

toplot <- svysmoothed$smooth
mapPlot(data=toplot, geo=et_admin1_map, variables=c("mean.original"),
        labels=c("Posterior Mean"), by.data="region",
        by.geo = "REG_P_CODE")

```



```{r direct_est}

toplot_direct <- svysmoothed$HT
mapPlot(data=toplot_direct, geo=et_admin1_map, variables=c("HT.est.original"),
        labels=c("Direct Estimates"), by.data="region",
        by.geo = "REG_P_CODE")

```


```{r smoothing_zones}

svysmoothed.zone <- fitGeneric(data=et_w3_data, geo=admin2_map,
                               Amat = mat.2, responseType = "binary", responseVar = "stunted",
                               strataVar = NULL, weightVar = "pw_w3", regionVar = "zone.st",
                               clusterVar = "ea_id2", CI=0.95)

```




```{r trying_a_different_way, eval=FALSE}

layerContrib=merge(et_admin1_map,toplot_direct, by.x='REG_P_CODE', by.y='region',all.x=F)
library(RColorBrewer)
library(classInt)
varToPLot=layerContrib$HT.est.original
numberOfClasses = 5
colorForScale='YlGnBu'
colors = brewer.pal(numberOfClasses, colorForScale)
intervals <- classIntervals(varToPLot, numberOfClasses, 
                            style = "quantile",
                            dataPrecision=2)
colorPallette <- findColours(intervals, colors)

legendText="intervals"
shrinkLegend=0.4
title="Proportion children u5 stunted in Ethiopia in 2015"

# first the ORIGINAL to signal missing values:
plot(et_admin1_map,col='red',main=title, lwd=3) 

# now the info on contributions
plot(layerContrib, lwd=3, col = colorPallette,border=NA,add=T) #add

# this uses all previous information
legend('topright', 
       legend = names(attr(colorPallette, "table")), #values
       fill = attr(colorPallette, "palette"), #colors
       cex = shrinkLegend, #size 
       bty = "n", # no box
       title=legendText)


```



### 4. Spatial smoothing models

```{r , eval=FALSE}

pcprec <- list(theta=list(prior="pc.prec", param=c(1, 0.01)))
ethiopia_fit1 <- inla(stunted ~ 1 + f(region, model="iid"), data=direct_region, family="logistic", control.predictor = list(compute=TRUE), verbose=TRUE)

```



### 5. Spatial point data

Start by estimating the temperature over the country

```{r krig_setup}

library(doBy)
library(geoR)

#setting up geovars
et_w3_geovars$log_precip <- log(et_w3_geovars$h2015_tot)

et_w3_geovars <- et_w3_geovars %>% 
  rename(
    ann_precip=h2015_tot
  )
et_w3_geovars$ann_temp <- et_w3_geovars$af_bio_1/10

et_w3_geovars.df <- as.data.frame(et_w3_geovars)

#collapsing geovars by cluster - they should be the same at each cluster
et_w3_geodat <- summaryBy(ann_precip + log_precip + ann_temp ~ lat_dd_mod + lon_dd_mod, data=et_w3_geovars.df, FUN=c(mean))

geo_mat <- matrix(cbind(et_w3_geodat$lat_dd_mod, et_w3_geodat$lon_dd_mod, et_w3_geodat$ann_temp, et_w3_geodat$log_precip.mean, et_w3_geodat$ann_temp),
                  ncol=5, nrow=530)

geo_pre <- as.geodata(geo_mat, coords.col = c(2, 1), data.col = c(3), covar.col=c(4, 5))

```


```{r krig_setup_stunting}

#set up geodata
et_w3_stunt <- as.data.frame(contriET_geo)
et_w3_stunt <- merge(et_w3_stunt, et_w3_geovars, no.dups=TRUE) 

et_w3_stunt.1 <- summaryBy(stunted + ann_precip + ann_temp ~ latitude + longitude, data=et_w3_stunt, FUN=c(mean))

et_w3_stunt_geo <- as.geodata(et_w3_stunt.1, coords.col = c("latitude", "longitude"), data.col = c("stunted.mean"), covar.col = c("ann_temp.mean", "ann_precip.mean"))

```



```{r krig_setup_wasting}

#set up geodata
et_w3_waste <- as.data.frame(contriET_geo)
et_w3_waste <- merge(et_w3_waste, et_w3_geovars, no.dups=TRUE)

et_w3_waste.1 <- summaryBy(wasted + ann_precip + ann_temp ~ latitude + longitude, data=et_w3_waste, FUN=c(mean))

et_w3_waste_geo <- as.geodata(et_w3_waste.1, coords.col = c("latitude", "longitude"), data.col = c("wasted.mean"), covar.col = c("ann_temp.mean", "ann_precip.mean"))


```



```{r variogram}

var_cloud <- variog(geo_pre, option="cloud")
plot(var_cloud, col="blue", cex=0.4)

#binned variogram with no trend
var_bin <- variog(geo_pre, uvec=seq(0, 12, 0.5))
plot(var_bin, cex=0.5, col="blue")

#testing with MC
precip.env <- variog.mc.env(geo_pre, obj=var_bin)
plot(var_bin, env=precip.env, col="red", ylim=c(0,30))

```


```{r stunted_variograms}

##With no covariates
#cloud variogram
stunt_cloud <- variog(et_w3_stunt_geo, option="cloud")
plot(stunt_cloud, col="blue", cex=0.4)

#binned variogram
stunt_bin <- variog(et_w3_stunt_geo, uvec=seq(0, 12, 0.5))
plot(stunt_bin, cex=0.5, col="blue")

#testing with MC
stunt.env <- variog.mc.env(et_w3_stunt_geo, obj=stunt_bin)
plot(stunt_bin, env=stunt.env, col="red")


##Adjusting for precip and temp
#bin
stunt_bin.1 <- variog(et_w3_stunt_geo, uvec=seq(0, 12, 0.5), trend = ~et_w3_stunt.1$ann_precip.mean + et_w3_stunt.1$ann_temp.mean)
plot(stunt_bin.1, cex=0.5, col="blue")

#testing with MC
stunt.env.1 <- variog.mc.env(et_w3_stunt_geo, obj=stunt_bin.1)
plot(stunt_bin.1, env=stunt.env.1, col="red")

```


```{r wasted_variograms}

##With no covariates
#cloud variogram
waste_cloud <- variog(et_w3_waste_geo, option="cloud")
plot(waste_cloud, col="blue", cex=0.4)

#binned variogram
waste_bin <- variog(et_w3_waste_geo, uvec=seq(0, 12, 0.5))
plot(waste_bin, cex=0.5, col="blue")

#testing with MC
waste.env <- variog.mc.env(et_w3_waste_geo, obj=waste_bin)
plot(waste_bin, env=waste.env, col="red")


##Adjusting for precip and temp
#bin
waste_bin.1 <- variog(et_w3_waste_geo, uvec=seq(0, 12, 0.5), trend = ~et_w3_waste.1$ann_precip.mean + et_w3_waste.1$ann_temp.mean)
plot(waste_bin.1, cex=0.5, col="blue")

#testing with MC
stunt.env.1 <- variog.mc.env(et_w3_stunt_geo, obj=stunt_bin.1)
plot(stunt_bin.1, env=stunt.env.1, col="red")

```


```{r fit_models}

theta <- -4/log(0.05)
  
olsfit <- variofit(var_bin, ini=c(15, theta), weights = "equal")
wlsfit <- variofit(var_bin, ini = c(15, theta))
mlfit <- likfit(geo_pre, ini = c(15, theta))
remlfit <- likfit(geo_pre, ini=c(15, theta), lik.method="RML")

plot(var_bin, max.dist=8, xlab = "Distance (km)", ylab="Semi-variance", pch=19, cex=0.6)
lines(olsfit, max.dist=8, col="red")
lines(wlsfit, max.dist=8, lty=2, col="green")
lines(mlfit, max.dist=8, lty=3, col="blue")
lines(remlfit, max.dist=8, lty=4, col="black")
legend("bottomright", legend=c("OLS", "WLS", "ML", "REML"), lty=c(1,2,3,4), bty="n", col=c("red", "green", "blue", "black"), cex=0.5)

```


```{r fit_models_2}

theta = -4/log(0.05)

olsfit.1 <- variofit(stunt_bin, ini=c(0.1, theta), weights="equal")
wlsfit.1 <- variofit(stunt_bin, ini = c(0.1, theta))
mlfit.1 <- likfit(et_w3_stunt_geo, ini = c(0.1, theta))
remlfit.1 <- likfit(et_w3_stunt_geo, ini=c(0.1, theta), lik.method="RML")

plot(stunt_bin, max.dist=10, xlab = "Distance (km)", ylab="Semi-variance", pch=19, cex=0.6)
lines(olsfit.1, max.dist=10, col="red")
lines(wlsfit.1, max.dist=10, lty=2, col="green")
lines(mlfit.1, max.dist=10, lty=3, col="blue")
lines(remlfit.1, max.dist=10, lty=4, col="black")
legend("bottomright", legend=c("OLS", "WLS", "ML", "REML"), lty=c(1,2,3,4), bty="n", col=c("red", "green", "blue", "black"), cex=0.5)

```


```{r fit_models_3}

theta = -4/log(0.05)

olsfit.2 <- variofit(waste_bin, ini=c(0.01, theta), weights="equal")
wlsfit.2 <- variofit(waste_bin, ini = c(0.01, theta))
mlfit.2 <- likfit(et_w3_waste_geo, ini = c(0.01, theta))
remlfit.2 <- likfit(et_w3_waste_geo, ini=c(0.01, theta), lik.method="RML")

plot(waste_bin, max.dist=10, xlab = "Distance (km)", ylab="Semi-variance", pch=19, cex=0.6)
lines(olsfit.2, max.dist=10, col="red")
lines(wlsfit.2, max.dist=10, lty=2, col="green")
lines(mlfit.2, max.dist=10, lty=3, col="blue")
lines(remlfit.2, max.dist=10, lty=4, col="black")
legend("bottomright", legend=c("OLS", "WLS", "ML", "REML"), lty=c(1,2,3,4), bty="n", col=c("red", "green", "blue", "black"), cex=0.5)

```


```{r}

lmfit <- lm(geo_pre$data ~ 1)
detrend <- as.geodata(cbind(geo_pre$coords, lmfit$residuals))

var_bin.2 <- variog(detrend, uvec=seq(0, 12, 0.5))
olsfit2 <- variofit(var_bin.2, ini=c(15, theta), weights="equal")

pred.grid <- expand.grid(seq(33.47,43.88, l=101), seq(3.45, 14.46, l=101))
kc <- krige.conv(detrend, loc=pred.grid,
                 krige=krige.control(obj.m=olsfit2))

```

```{r}

lmfit.1 <- lm(et_w3_stunt_geo$data ~ 1)
detrend.1 <- as.geodata(cbind(et_w3_stunt_geo$coords, lmfit.1$residuals))

mlfit.2 <- likfit(detrend.1, ini=c(0.1, theta))

pred.grid.1 <- expand.grid(seq(33.47,43.88, l=101), seq(3.45, 14.46, l=101))
kc.1 <- krige.conv(detrend.1, loc=pred.grid.1,
                   krige=krige.control(obj.m=mlfit.2))

```


```{r}

lmfit.2 <- lm(et_w3_waste_geo$data ~ 1)
detrend.2 <- as.geodata(cbind(et_w3_waste_geo$coords, lmfit.2$residuals))

waste_bin.2 <- variog(detrend, uvec=seq(0,12,0.5))
wlsfit.2 <- variofit(waste_bin.2, ini = c(0.01, theta))

pred.grid.2 <- expand.grid(seq(33.47,43.88, l=101), seq(3.45, 14.46, l=101))
kc.2 <- krige.conv(detrend.2, loc=pred.grid.2,
                   krige=krige.control(obj.m=wlsfit.2))

```



```{r krige_image}

library(fields)
image.plot(x=pred.grid[["Var1"]][1:101], y=unique(pred.grid[["Var2"]]),
           z=matrix(kc$predict, nrow=101, ncol=101), col=terrain.colors(100),
           xlab="longitude", ylab="latitude")
symbols(detrend$coords[,1], detrend$coords[,2], circles=(detrend$data-min(detrend$data))/1, add=T, inches=0.04)
plot(et_admin1_map,border='black',lwd=1, add=T)


```

```{r}

image.plot(x=pred.grid.1[["Var1"]][1:101], y=unique(pred.grid.1[["Var2"]]),
           z=matrix(kc.1$predict, nrow=101, ncol=101), col=terrain.colors(100),
           xlab="longitude", ylab="latitude")
symbols(detrend.1$coords[,2], detrend.1$coords[,1], circles=(detrend.1$data-min(detrend.1$data))/1, add=T, inches=0.04)
plot(et_admin1_map,border='black',lwd=1, add=T)


```


```{r}

image.plot(x=pred.grid.2[["Var1"]][1:101], y=unique(pred.grid.2[["Var2"]]),
           z=matrix(kc.2$predict, nrow=101, ncol=101), col=terrain.colors(100),
           xlab="longitude", ylab="latitude")
symbols(detrend.2$coords[,1], detrend.2$coords[,2], circles=(detrend.2$data-min(detrend.2$data))/1, add=T, inches=0.04)
plot(et_admin1_map,border='black',lwd=1, add=T)


```


### 6. Regression analysis


```{r map_stunting}

spplot(direct_region)

```


```{r inla}

inla.region.1 <- inla(stunted ~ f(region, model="iid"),
                       family = "poisson", E = direct_region$stunted, data=direct_region,
                       control.predictor=list(compute = TRUE))



```




